---
title: "Global Language Diversity on Wikipedia"
description: "A table showing the relationship between language speaker populations and Wikipedia article counts."
author: "Carwil Bjork-James"
date: "February 23, 2026"
format: 
  html:
    self-contained: true
    page-layout: full
    toc: true
---

# Languages and Wikipedia Articles

![The largest Wikipedia editions by number of articles, 2024 (CC-0 [Arief Azazie Zain](https://commons.wikimedia.org/wiki/User:Arief_Azazie_Zain "User:Arief Azazie Zain")) ](/images/wikipedia/List_of_Wikipedia_articles_by_language_-_March_2024.svg.png){alt="The largest Wikipedia editions by number of articles"}

```{r setup, include=FALSE}
library(tidyverse)
library(rvest)
library(knitr)
library(kableExtra)
library(stringi)
library(viridis)
library(scales)

linguameta_root_path <- "~/Dropbox (Personal)/R/url-nlp/linguameta"
```

```{r load-data, echo=FALSE, message=FALSE, warning=FALSE}
# ----------------------------------------------------------------------
# 1. Load linguameta
# ----------------------------------------------------------------------
linguameta <- read_tsv(
  file = file.path(linguameta_root_path, "linguameta.tsv"),
  show_col_types = FALSE
)

# ----------------------------------------------------------------------
# 2. Scrape live Wikipedia table and clean column names
# ----------------------------------------------------------------------
wiki_html <- read_html("https://en.wikipedia.org/wiki/List_of_Wikipedias")

wiki <- wiki_html |>
  html_table() |>
  keep(~ "Articles" %in% names(.)) |>
  _[[1]] |>
  rename(
    wiki_name_en     = `Wikipedia name in English`,
    wiki_name_native = `Wikipedia namein native language`,
    language         = Language,
    script           = `Script (ISO15924 code)`,
    wp_code          = WPcode,
    articles         = Articles,
    active_users     = Activeusers,
    launch_date      = `Launch date`
  ) |>
  mutate(
    articles     = as.numeric(gsub(",", "", articles)),
    active_users = as.numeric(gsub(",", "", active_users))
  ) |>
  filter(articles > 0)

# ----------------------------------------------------------------------
# 3. Join wiki → linguameta via BCP-47 code
#    Some Wikipedia edition codes are non-standard (zh-yue, zh-min-nan,
#    or carry "(closed)" suffixes); map them to canonical BCP-47 tags.
# ----------------------------------------------------------------------
wp_to_bcp47_extra <- c(
  "zh-yue"     = "yue",   # Cantonese
  "zh-min-nan" = "nan",   # Southern Min
  "cr"         = "cr",    # Cree
  "kl"         = "kl",    # Kalaallisut / Greenlandic
  "ak"         = "ak",    # Akan / Twi
  "kr"         = "kr",    # Kanuri
  "aa"         = "aa",    # Afar
  "ii"         = "ii",    # Sichuan Yi
  "ho"         = "ho",    # Hiri Motu
  "hz"         = "hz",    # Herero
  "kj"         = "kj",    # Kuanyama
  "mh"         = "mh",    # Marshallese
  "na"         = "na",    # Nauruan
  "ng"         = "ng",    # Ndonga
  "lrc"        = "lrc"    # Northern Luri
)

wiki_joined <- wiki |>
  mutate(
    # strip " (closed)", " (deleted)" etc.
    wp_code_clean = str_trim(str_remove(wp_code, "\\s*\\(.*")),
    bcp47_lookup  = coalesce(wp_to_bcp47_extra[wp_code_clean], wp_code_clean)
  ) |>
  select(bcp47_lookup, wp_code, articles, active_users, language)

# Wiki editions that don't map to any spoken language in linguameta
# (constructed languages, classical languages, Simple English, dialects
#  with non-standard codes, etc.)
wiki_special <- wiki |>
  mutate(
    wp_code_clean = str_trim(str_remove(wp_code, "\\s*\\(.*")),
    bcp47_lookup  = coalesce(wp_to_bcp47_extra[wp_code_clean], wp_code_clean)
  ) |>
  filter(!bcp47_lookup %in% linguameta$bcp_47_code) |>
  select(language, wp_code, articles, active_users)

text_to_num <- function(text) {
  text <- str_replace_all(text, ",", "")
  multipliers <- c(k = 1e3, K = 1e3, M = 1e6, B = 1e9)
  pattern <- "([0-9.]+)\\s*([KMBk]?)"
  
  as.numeric(sapply(text, function(x) {
    matches <- str_match(x, pattern)
    if (is.na(matches[1,1])) {
      return(NA)
    }
    number <- as.numeric(matches[1,2])
    suffix <- matches[1,3]
    if (suffix != "") {
      number * multipliers[[suffix]]
    } else {
      number
    }
  }))
}

# ----------------------------------------------------------------------
# 4. Supplementary speaker estimates for languages where linguameta
#    has NA (sourced from Wikipedia language articles)
# ----------------------------------------------------------------------
na_speakers_supplement <- tribble(
  ~bcp_47_code, ~speakers_supplement, ~speakers_source,
  "tl",   33000000,  "L1 speakers (Ethnologue 2023, via Wikipedia)",
  "crh",    580000,  "Native speakers (Ethnologue 2001, via Wikipedia)",
  "nrm",    100000,  "Approximate (Norman language, France/Channel Islands)",
  "bxr",    218000,  "Russia Buryat (Ethnologue estimate)",
  "szy",     10000,  "Sakizaya (estimated, Taiwan)",
  "rsk",    600000,  "Pannonian Rusyn/Ruthenian (approximate)",
  "bdr",    200000,  "West Coast Bajau (approximate)"
)

# ----------------------------------------------------------------------
# 5. Attach Wikipedia data to linguameta; fill NA speakers; assign
#    global rank; then partition into three groups:
#      A. Top 250 by speakers
#      B. Smaller languages with a Wikipedia (outside top 250)
#      C. Special / constructed / classical Wikipedias (no linguameta row)
# ----------------------------------------------------------------------
linguameta_with_wiki <- linguameta |>
  left_join(wiki_joined, by = c("bcp_47_code" = "bcp47_lookup")) |>
  left_join(na_speakers_supplement, by = "bcp_47_code") |>
  mutate(
    speakers = coalesce(estimated_number_of_speakers, speakers_supplement)
  ) |>
  mutate(
    global_rank = row_number(desc(coalesce(speakers, 0L)))
  ) |>
  filter(global_rank <= 250 | !is.na(articles)) |>
  mutate(articles = replace_na(articles, 0))

# Partition
top250   <- linguameta_with_wiki |> 
  filter(global_rank <= 250) |> 
  arrange(desc(speakers))
extras   <- linguameta_with_wiki |> 
  filter(global_rank > 250, articles > 0) |> 
  arrange(desc(speakers))

# ----------------------------------------------------------------------
# 6. Helper: format a data frame for display
# ----------------------------------------------------------------------
format_for_display <- function(df, show_rank = TRUE) {
  df |>
    mutate(
      rank_label = if (show_rank) as.character(global_rank) else "—",
      is_roman   = stri_detect_regex(
        endonym, "^[\\p{Script=Latin}[:punct:][:space:]0-9]+$"
      ),
      autonym_fmt = if_else(
        !is.na(endonym) & is_roman,
        paste0(" <i>", endonym, "</i>"),
        paste0(" ", replace_na(endonym, ""))
      ),
      language_display = paste0(rank_label, ". ", english_name, "<br>", autonym_fmt),
      articles_fmt     = format(articles, big.mark = ",")
    )
}

# ----------------------------------------------------------------------
# 7. Colour scale helpers (shared domain across both tables)
# ----------------------------------------------------------------------
all_speakers <- linguameta_with_wiki$speakers
color_pal <- col_numeric(
  palette = c("#ffffff", "#74a9cf", "#0570b0"),
  domain  = log1p(c(all_speakers, 1000000, NA))
)

make_article_colors <- function(articles) {
  case_when(
    articles == 0          ~ "#ffffff",
    articles < 10000       ~ "#e6f7f0",
    articles < 100000      ~ "#bae4d7",
    articles < 1000000     ~ "#82c9a5",
    articles < 4000000     ~ "#c1d383",
    articles >= 4000000    ~ "#ffd700"
  )
}

render_table <- function(df) {
  fmt <- format_for_display(df, show_rank = TRUE)
  speaker_colors <- color_pal(log1p(fmt$speakers))
  article_colors <- make_article_colors(fmt$articles)

  fmt |>
    select(language_display, wikidata_description, speakers, articles_fmt) |>
    kable(
      format    = "html",
      escape    = FALSE,
      col.names = c("Language", "Language Description",
                    "Speakers (L1 est.)", "Wikipedia Articles"),
      align     = "llrr"
    ) |>
    kable_styling(
      bootstrap_options = c("hover", "condensed"),
      font_size         = 14,
      htmltable_class   = "table table-hover table-condensed"
    ) |>
    column_spec(1, width = "35%") |>
    column_spec(2, width = "35%") |>
    column_spec(3, background = speaker_colors, width = "15%") |>
    column_spec(4, background = article_colors, width = "15%") |>
    row_spec(0, background = "white", color = "black", bold = TRUE)
}
```

## Top 250 Languages by Number of Speakers

Speaker estimates are L1 (first-language) counts from [LinguaMeta](https://github.com/cisnlp/linguameta), supplemented for a handful of languages where LinguaMeta lacks an estimate (sourced from Wikipedia). Background shading: **blue** = speaker population; **green/gold** = Wikipedia article count.

:::{.column-page-inset-right}
```{r table-top250, echo=FALSE, message=FALSE, warning=FALSE}
render_table(top250)
```
:::

## Additional Languages with Wikipedia Editions

These languages fall outside the top 250 by speaker population but have active Wikipedia editions. Global rank (by estimated L1 speakers) shown where available.

:::{.column-page-inset-right}
```{r table-extras, echo=FALSE, message=FALSE, warning=FALSE}
extras_fmt <- format_for_display(extras, show_rank = TRUE)
speaker_colors_e <- color_pal(log1p(extras_fmt$speakers))
article_colors_e <- make_article_colors(extras_fmt$articles)

extras_fmt |>
  select(language_display, wikidata_description, speakers, articles_fmt) |> 
  kable(
    format    = "html",
    escape    = FALSE,
    col.names = c("Language", "Language Description",
                  "Speakers (L1 est.)", "Wikipedia Articles"),
    align     = "llrr"
  ) |>
  kable_styling(
    bootstrap_options = c("hover", "condensed"),
    font_size         = 14,
    htmltable_class   = "table table-hover table-condensed"
  ) |>
  column_spec(1, width = "35%") |>
  column_spec(2, width = "35%") |>
  column_spec(3, background = speaker_colors_e, width = "15%") |>
  column_spec(4, background = article_colors_e, width = "15%") |>
  row_spec(0, background = "white", color = "black", bold = TRUE)
```
:::

## Classical, Constructed, and Special-Purpose Wikipedia Editions

These Wikipedia editions are not associated with a living natural language in the LinguaMeta dataset. They include classical and liturgical languages (Latin, Sanskrit, Old English), constructed languages (Esperanto is covered above; Ido, Interlingua, Volapük, etc. appear here), dialect/regional-code editions, and the Simple English Wikipedia.

:::{.column-page-inset-right}
```{r table-special, echo=FALSE, message=FALSE, warning=FALSE}
article_colors_s <- make_article_colors(wiki_special$articles)

wiki_special |>
  mutate(articles_fmt = format(articles, big.mark = ",")) |>
  arrange(desc(articles)) |>
  select(language, wp_code, articles_fmt) |>
  kable(
    format    = "html",
    escape    = FALSE,
    col.names = c("Wikipedia Edition", "WP Code", "Articles"),
    align     = "llr"
  ) |>
  kable_styling(
    bootstrap_options = c("hover", "condensed"),
    font_size         = 14,
    htmltable_class   = "table table-hover table-condensed"
  ) |>
  column_spec(1, width = "50%") |>
  column_spec(2, width = "20%") |>
  column_spec(3, background = article_colors_s, width = "30%") |>
  row_spec(0, background = "white", color = "black", bold = TRUE)
```
:::
